---
title: "Investigation on data-related jobs across United States"
author: "Jiaxu Luo"
date: "2016/07/18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Summary

### When a U.S. company wants to hire someone from outside of the United States for a special tier or technical position, they have to file an application with the United States government to get a visa for the applicant so that the applicant is allowed to stay in the country. Typically, the company will either file a green card application, a labor condition application, called an LCA. workers. And each time a company files an application, they have to state how much they are planning on paying the employee. They also have to state the average amount an employee with similar skills and background typically gets paid for the same position.
### The dataset is downloaded at the Office of Foreign Labor Certification website. 
### It would be possible to look up information about what types of salaries people in a particular profession are making. It would also be possible to look up how these wages vary across different parts of the country. Since many of you taking the class might be considering making a career move into a data-related field, I thought it might be interesting to use this publicly available data to explore what types of salaries you might encounter.


## Question1: what is the data-related jobs demand distribution across United States?

### Here, I regard the job title "data scientist","business analyst","business analyst" as data-related jobs.
```{r}
## Load Libraries and read data

library(dplyr)
library(ggplot2)
library(maps)
library(ggmap)

SalaryData=read.csv("Salary.csv",stringsAsFactors = F)

## Getting the lat/lon of the State of America (excluding Alaska & Hawaii),DC included
DC=data.frame(state.name="District of Columbia",Abr="DC")
Gcode=read.csv("Gcode.csv")
States= data.frame(state.name,Abr=state.abb) %>%
        rbind(DC) %>%
        filter(!state.name %in% c("Alaska","Hawaii")) %>%
        cbind(Gcode)
States[45,3]=-120
States[45,4]=47

# Question1: what is the data-related jobs demand distribution across United States?
##1.1 Create a table for demand mapping

SalaryData_Join=inner_join(SalaryData,States,by=c("WORK_STATE"="Abr"))

Number_Of_Demand= filter(SalaryData_Join,JOB_TITLE_SUBGROUP %in% c("data scientist","data analyst","business analyst")) %>%
        group_by(WORK_STATE) %>%
        count(WORK_STATE)

Number_Of_Demand=select(Number_Of_Demand,WORK_STATE=WORK_STATE,Number=n)
Demand_table=inner_join(Number_Of_Demand,States,by=c("WORK_STATE"="Abr"))


##1.2 mapping the demand on USA_State Map using size
Demand_table$Number=as.numeric(Demand_table$Number)
usa_center = as.numeric(geocode("United States"))
USAMap = ggmap(get_googlemap(center=usa_center, scale=2, zoom=4), extent="panel")
graph1=USAMap +
        geom_point(aes(x=lon, y=lat,size=Number), data=Demand_table, col="blue", alpha=0.4,show.legend = T) +
        scale_size_area(max_size = 20)+
        ggtitle("Demand on Data-related jobs across USA(sample)")
plot(graph1)
```

### It turns out that California, New York, New Jersy and Texas have the highest demand on data-related jobs.

## Question2: which city in CA,NY,NJ,TX have more demand on data-related jobs?
```{r}
## Question2: which city in CA,NY,NJ,TX have more demand on data-related jobs?
Cities= filter(SalaryData,WORK_STATE %in% c("NY","NJ","CA","TX")) %>%
        filter(JOB_TITLE_SUBGROUP %in% c("data scientist","data analyst","business analyst")) %>%
        group_by(WORK_CITY) %>%
        count(WORK_CITY) %>%
        arrange(desc(n)) %>%
        slice(1:8)

g=ggplot(Cities,aes(reorder(x=WORK_CITY,n,desc),y=n))      
g+geom_bar(stat = "identity",aes(fill=WORK_CITY))+
        ggtitle("Demand on data-related job across cities")+
        xlab("Cities")

```


## Question3 :which states provides higher average salary for data-related jobs?
```{r}
# Question3 :which states provides higher average salary for data-related jobs?
##3.1 Create a table for salary mapping
Salary_state= filter(SalaryData_Join,JOB_TITLE_SUBGROUP %in% c("data scientist","data analyst","business analyst")) %>%
        group_by(WORK_STATE) %>%
        summarise(Avg_pay=mean(PAID_WAGE_PER_YEAR))
Salary_table=inner_join(Salary_state,States,by=c("WORK_STATE"="Abr"))
Salary_table=Salary_table[,-c(5,6)]



##3.2 mapping the salary on USA_State Map using continuous color
Salary_table$state.name=tolower(Salary_table$state.name)
mapdata=map_data("state")
Salary_table_join=inner_join(Salary_table,mapdata,by=c("state.name"="region"))
g1=ggplot(data = Salary_table_join,aes(x=long,y=lat))
graph2=g1+geom_polygon(aes(color="darkblue",group = group,fill=Avg_pay),show.legend = F) +
        scale_fill_continuous(low = "orange", high = "darkblue")+
        ggtitle("Average Salary for Data-related job Across USA(sample)")
plot(graph2)

```

### It turns out that Washington, California and New York State pay the highest salary for data-related jobs.

## Question4  Which companys in CA,NY or WA provide more salary than others?
```{r}
##4.1 choose those companies that hire more than 8 employees
High_Pay_Comp= filter(SalaryData,WORK_STATE %in% c("NY","WA","CA")) %>%
        filter(JOB_TITLE_SUBGROUP %in% c("data scientist","data analyst","business analyst")) %>%
        group_by(EMPLOYER_NAME) %>%
        count(EMPLOYER_NAME) %>%
        filter(n>=8) 
Salary_Rank=filter(SalaryData,WORK_STATE %in% c("NY","WA","CA")) %>%
        filter(JOB_TITLE_SUBGROUP %in% c("data scientist","data analyst","business analyst")) %>%
        inner_join(High_Pay_Comp) %>%
        group_by(EMPLOYER_NAME) %>%
        summarise(AVG_salary=mean(PAID_WAGE_PER_YEAR)) %>%
         arrange(desc(AVG_salary)) %>%
        slice(1:10)


##4.2  List those companies with highest salaries
g2=ggplot(Salary_Rank,aes(reorder(x=EMPLOYER_NAME,AVG_salary,desc),y=AVG_salary))      
       g2+geom_bar(stat = "identity",aes(fill=EMPLOYER_NAME))+
               ggtitle("Companies pay highest salary")+
               xlab("Companys")
```