---
title: "Anslysis of the Rent The Runway"
author: "Yunrou Gong (Crystal)"
date: "August 14, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Introduction:   
- Backgroud Review   
- Methodology  
- Visiualization the data  
-  
-  
```{r loadingdata, include=FALSE}
library(dplyr)
library(googleVis)
library(ggplot2)
rwdata<- read.csv("data.csv")
rwdata$item <- as.character(rwdata$item)
```

## Summary of Data 


```{r pressure, echo=FALSE}
str(rwdata)
summary(rwdata[,-8])


```
## Exploratory Data Analysis
### Categorical Variable : Brand and Type 

```{r brand eda, echo=FALSE, results="asis"}
brand_share <- count(rwdata, vars=brand,sort = T)
Pie.brand <- gvisPieChart(brand_share, options=list(
  slices="{14: {offset: 0.2}, 0: {offset: 0.3}}",
  title='Brand Share in Pie Chart',
  legend="{ position:'left'}",
  width=800, height=600,
  pieSliceText='label',
  pieHole=0.5))
plot(Pie.brand,"chart")

type_share <- count(rwdata, vars=type,sort = T)
Pie.type <- gvisPieChart(type_share, options=list(
  slices="{6: {offset: 0.2}, 1: {offset: 0.3}}",
  title='Dress Type Share in Pie Chart',
  legend="{ position:'left'}",
  width=800, height=600,
  pieSliceText='label',
  pieHole=0.5))
plot(Pie.type,"chart")

```

##Popularity by dress and by brand 
```{r popularity, echo=FALSE, results="asis"}

rwdata$item <- as.character(rwdata$item)

top20 <- head(rwdata%>%arrange(desc(reviews)),20)

g.bar <- gvisColumnChart(top20,xvar ="item", yvar = "reviews",
                         options=list(color="#99ccff",
                                      width="1000px", height="500px",
                                      vAxes="[{title:'Number of Reviews'}]",
                                      hAxes="[{title:'Dress Name'}]",
                                      title="Top 20 Popular Dresses ",
                                      titleTextStyle="{fontSize:18}",
                                      legend="{position:'Top'}"))
plot(g.bar,"chart")



rwdata$brand <- as.character(rwdata$brand)
brand_bar <- rwdata%>%group_by(brand)%>%
  summarise(reviews_brand=sum(reviews))%>%
  arrange(desc(reviews_brand))

brand_bar <- brand_bar[1:20,]
brand.barchart <- gvisColumnChart(brand_bar,xvar ="brand", yvar = "reviews_brand",
                         options=list(colors="#cc99ff",
                                      width="1000px", height="500px",
                                      vAxes="[{title:'Aggregate number of Reviews'}]",
                                      hAxes="[{title:'Brand Name'}]",
                                      title="Top 20 Popular Brands ",
                                      titleTextStyle="{fontSize:18}",
                                      legend="{position:'Top'}"))
plot(brand.barchart,"chart")




```

#
```{r scatter, echo=FALSE, results="asis"}

mydata <- rwdata[-2]
mydata <- mydata[-7]
mydata <- mydata[-10]
pairs(mydata[-1])


```

# EDA of numeric variable 
```{r hist, echo=FALSE, results="asis"}

hist(x=rwdata$avgRating,prob=T)

hist(x=rwdata$discount,prob=T)
lines(density(rwdata$discount), col = "purple")
hist(x=rwdata$retailPrice,prob=T)
lines(density(rwdata$retailPrice), col = "orange")
hist(x=rwdata$rentalPrice_4,prob=T)
lines(density(rwdata$rentalPrice_4), col = "red")

```

###data by selected relevevant variables to predict rental price 
response variable = rentalPrice_4
```{r sampledata, include=FALSE, results="asis"}
library(ISLR)
library(glmnet)
mydata2 <- rwdata%>%select(item,brand,type,rentalPrice_4,retailPrice)
mydata2$brand <- as.factor(mydata2$brand)
library(car)
scatterplotMatrix(~ rentalPrice_4 + retailPrice | type, data=mydata2,
                  main="Scaterplot of rentalPrice_4 and retailPrice")

# set.seed(0)
# x <- model.matrix(rentalPrice_4 ~ retailPrice, mydata2)[, -1] 
# y <- mydata2$rentalPrice_4
# train.rw<-  sample(1:length(x), 8*length(x)/10)
# 
# z <-as.matrix(mydata2[-4])
# train.zrw<-  sample(1:nrow(z), 8*nrow(z)/10)


```
# scatter plot pink

```{r scatter2, include=FALSE, results="asis"}

s <- gvisScatterChart(rwdata[,c(10,11)],
                      options=list(colors="pink",
                                   width="1000px", height="500px",
                                   vAxes="[{title:'rental Price of 8 days'}]",
                                   hAxes="[{title:'rental Price of 4 days'}]",
                                   title="ScatterPlot rentalPrice_8 Vs rentalPrice_4",
                                   titleTextStyle="{fontSize:18}",
                                   legend="{position:'Top'}"))

plot(s,"chart")

s2 <- gvisScatterChart(rwdata[,c(9,10)],
                      options=list(colors="hotpink",
                                   width="1000px", height="500px",
                                   vAxes="[{title:'rental Price of 4 days'}]",
                                   hAxes="[{title:'retail price of dress'}]",
                                   title="ScatterPlot rentalPrice_4 Vs retailPrice",
                                   titleTextStyle="{fontSize:18}",
                                   legend="{position:'Top'}"))


plot(s2,"chart")




```


```{r model, echo=FALSE, results="asis"}
##model-mutiple linear regression 
model1 <- lm(rentalPrice_4 ~ . -item, data=mydata2)
summary(model1)
model2 <- lm(rentalPrice_4 ~ . -item -brand , data=mydata2)
summary(model2)
model3 <- lm(rentalPrice_4 ~ retailPrice,data=mydata2)
summary(model3)
plot(model2)
influencePlot(model3)
vif(model3)
avPlots(model3)

```
The overall regression is significant as the F -test P-value is less than 0.05 , thus there is at   least one of the variable in the model is potentially useful for prediction the Price .  



Adjusted coefficient of determination: 0.8479 
approximately 85% of the variability in the Price variable 
is explained by the model.


plot(model)
influencePlot(model)
vif(model)
avPlots(model)
```{r cpmparemodel, echo=FALSE, results="asis"}

AIC(model1,model2,model3)
BIC(model1,model2,model3)


```

result <- predict(model3, newdata=mydata2)

