---
title: "Bitcoin - Presentation 1"
author: "Ben Townson"
date: "July 18, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(Quandl)
library(quantmod)
library(dplyr)
library(ggplot2)
library(reshape)
library(scales)

bc_ohlc=readRDS("bc_ohlc.rds")
```

## Bitcoin - Does It Have Legs?
```{r leader, echo=F,fig.height=3,fig.width=3}
library(png)
library(grid)
img <- readPNG("bc_img.png")
 grid.raster(img)
```
  
* **Bitcoin is:**
* Electronic currency, exchanged via blockchain technology
* Traded currency, transacted on various exchanges

##The Datasets
* **Trading data from Bitcoincharts.com**
* Consolidated trades from all bitcoin exchanges
* includes daily high, low, open, and close data

* **Bitcoin blockchain transaction data from Blockchain.info**
* Total Transactions: Non-exchange transactions on the blockchain
* Unique Addresses: Unique bitcoin users, excluding top 100 addresses
* Transaction Volume: USD value of transactions in bitcoin

## Highly Volatile Currency
```{r vol1, echo=F}
g <- ggplot(bc_ohlc, aes(x=date, weighted.price)) 
g <- g + geom_line() + ggtitle("Daily Weighted Price for bitcoin")
g

```

* Maximum Daily Volatility (Annualized): `r paste0(round(100*max(bc_ohlc$realized.vol[20:length(bc_ohlc)]),2),'%')`
* Maximum Daily Move: `r max(bc_ohlc$range)`

## Volatility Highlighted
```{r vol2, echo=F}
g1 <- ggplot(bc_ohlc[732:854,], aes(x=date, lower=candleLower, middle = candleLower,upper=candleUpper,ymin=low,ymax=high)) 
g1 <- g1 + geom_boxplot(stat='identity', aes(group=date,fill=fill)) + ggtitle("Daily Price for bitcoin from November 2014-January 2015") + theme(legend.position = "none")
g1

```

## Volatility Highlighted
```{r vol3, echo=F,warning=F}

g2 <- ggplot(bc_ohlc,aes(return))
g2 <- g2 + geom_histogram(aes(fill=year),bins = 60) + ggtitle("Distribution of Daily Returns") + scale_x_continuous(labels=percent)
g2

```


## Adoption of Bitcoin
```{r adopt1, echo=F}

g3 <- ggplot(bc_ohlc,aes(date,unique.addresses))
g3 <- g3 + geom_point() + geom_smooth(method="loess") + ggtitle("Unique Bitcoin Addresses Used Daily") + scale_y_continuous(labels=comma)
g3

```

## Adoption of Bitcoin
```{r adopt2, echo=F}

g4 <- ggplot(bc_ohlc,aes(date,daily.transactions))
g4 <- g4 + geom_point() + geom_smooth(method="loess")+ ggtitle("Daily Estimated Transactions") + scale_y_continuous(labels=comma) 
g4

```

## Adoption of Bitcoin
```{r adopt3, echo=F}

g5 <- ggplot(bc_ohlc,aes(date,estimated.transaction.volume))
g5 <- g5 + geom_point() + geom_smooth(method="loess")+ ggtitle("Estimated USD Transaction Volume") + scale_y_continuous(labels=comma)
g5

```

## Unfazed by Volatility
```{r unfazed1, echo=F}

g6 <- ggplot(bc_ohlc,aes(weighted.price,daily.transactions))
g6 <- g6 + geom_point(aes(color=date)) + ggtitle("Weighted Price vs Daily Transactions") + scale_y_continuous(labels=comma)
g6

```

## Unfazed by Volatility
```{r unfazed2, echo=F,error=F}
g7 <- ggplot(bc_ohlc,aes(ret.bucket,estimated.transaction.volume))
g7 <- g7 + geom_bar(aes(fill=year),stat="identity",position='dodge')+ ggtitle("Transactions vs Daily Returns") + scale_y_continuous(labels=comma) + xlab("Return Buckets")
g7

```

## Unfazed by Volatility
```{r unfazed3, echo=F,error=F}
g8 <- ggplot(bc_ohlc,aes(range,daily.transactions))
g8 <- g8 + geom_point(aes(color=year))+ scale_y_continuous(labels=comma)+ ggtitle("Daily Price Range vs Daily Transactions")
g8

```

## Conclusion

- Users of bitcoin remain unfazed by volatility
- Growth of bitcoin as a means of transacting remains a function of time
- The future for bitcoin is not threatened by price and volatility