---
title: "We run faster than you do"
subtitle: "An analysis of marathon results from Switzerland"
author: "David Richard Steinmetz"
date: "August 14, 2016"
output:
  ioslides_presentation:
    widescreen: true
    #smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(data.table))
dt <- readRDS('data_clean.RDS')
nat.N <- head(dt[, .N, by = nat][order(-N)],10) 
nat <- subset(dt, nat %in% c('SUI', 'GER', 'NED', 'GBR', 'ITA'))
```

## Both external and personal factors affect your marathon times

Did you train hard enough?

You might consider this most when considering your marathon time and place, but a number of other factors are also at play. The number of participants, your gender, your age, can all have an affect, perhaps even your nationality.

In this presentation we will investigate these factors to see how they affected marathon times and placements in the Jungfrau Marathon in Switzerland.

## The year 2012 saw a lack of participants and longer race times

```{r}
# Boxplots
# --Year
g <- ggplot(dt, aes(as.factor(year), time))
g <- g + geom_boxplot(aes(colour=as.factor(year)))
g <- g + theme_minimal()
g <- g + ggtitle('Distribution of Race Times')
g <- g + xlab('Year') + ylab('Time (s)')
g <- g + scale_colour_discrete(name = "Year")
g
```


## Your gender appears to impact your time

```{r}
# Boxplots
# --Gender
g <- ggplot(dt, aes(as.factor(gender), time))
g <- g + geom_boxplot(aes(colour=as.factor(gender)))
g <- g + theme_minimal()
g <- g + ggtitle('Distribution of Race Times')
g <- g + xlab('Gender') + ylab('Time (s)')
g <- g + scale_colour_discrete(name = "Gender")
g
```

- T-test p-value < 2.2e-16
- BUT: T-TEST INVALID! An F-test shows the genders have differing variances


## Your age impacts your time

```{r}
# Boxplots
# --Age Group
g <- ggplot(dt[!is.na(age.cat)], aes(as.factor(age.cat), time))
g <- g + geom_boxplot(aes(colour=as.factor(age.cat)))
g <- g + theme_minimal()
g <- g + ggtitle('Distribution of Race Times')
g <- g + xlab('Age Group') + ylab('Time (s)')
g <- g + scale_colour_discrete(name = "Age Category")
g
```

## And your age also impacts your placement

```{r}
# --Age Group (Place)
g <- ggplot(dt[!is.na(age.cat)], aes(as.factor(age.cat), place))
g <- g + geom_boxplot(aes(colour=as.factor(age.cat)))
g <- g + theme_minimal()
g <- g + ggtitle('Distribution of Race Times')
g <- g + xlab('Age Group') + ylab('Place')
g <- g + scale_colour_discrete(name = "Age Category")
g
```

## Proximity to Switzerland influenced who participated

```{r}
# Barchart by nationality
g <- ggplot(nat.N, aes(nat, N))
g <- g + geom_bar(stat='identity')
g <- g + theme_minimal()
g <- g + ggtitle('Number of Participants by Nationality')
g <- g + xlab('Country') + ylab('Number')
g <- g + coord_flip()
g
```

## The Italians are the fastest

```{r}
# --Nationality reordered according to median time
g <- ggplot(nat, aes(reorder(as.factor(nat), time, median), time))
g <- g + geom_boxplot(aes(colour=as.factor(nat)))
g <- g + theme_minimal()
g <- g + ggtitle('Distribution of Race Times')
g <- g + xlab('Nationality (ordered by median time)') + ylab('Time (s)')
g <- g + scale_colour_discrete(name = "Nationality")
g
```

## Gender does not change time distribution by nationality 

```{r}
# --Gender + Nationality
g <- ggplot(nat, aes(as.factor(gender), time))
g <- g + geom_boxplot(aes(colour=reorder(nat, time)))
g <- g + theme_minimal()
g <- g + ggtitle('Distribution of Race Times')
g <- g + xlab('Gender') + ylab('Time (s)')
g <- g + scale_colour_discrete(name = "Nationality")
g
```

## Takeaway

- The Italians are the fastest
- Gender does not change nation rankings
- Age is running still matters: younger is better

