---
title: "Impact of HbA1c Measurement on Diabetic Medications and Hosiptal Readmission"
author: "Jingyu Zhang"
date: "July 17, 2016"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(dplyr)
library(ggplot2)
library(ggthemes)
library(RColorBrewer)

Diadata = read.csv(file = "diabetic_data.csv", stringsAsFactors = F)
# For Diadata1, fliter the data with missing race and gender
Diadata1 = Diadata[!((Diadata$race == "?")|(Diadata$gender == "Unknown/Invalid")), ]


Diadata2 <- group_by(Diadata, patient_nbr) %>%
  dplyr::summarise(., encounter_id = min(encounter_id)) %>%
  left_join(.,Diadata, by = "encounter_id") %>%
  select(., -1) %>%
  rename(., patient_nbr = patient_nbr.y) %>%
  filter(., discharge_disposition_id %in% c(1:10,12,15:17,18,19:29))
#For Diadata3, we filter the Diadata2 and remove the patient who is dischaged to hospice.
#The dimension is 69980 * 50.


fac = function(i) {
  if (i %in% c(390:459,785))  return('Circulatory')
  else if (i %in% c(460:519,786)) return('Respiratory')
  else if (i %in% c(520:579,787)) return('Digestive')
  else if (i>=250.0&i<251) return('Diabetes')
  else if (i %in% (800:999)) return('Injury')
  else if (i %in% (710:739)) return('Musculoskeletal')
  else if (i %in% c(580:629,788)) return('Genitourinary')
  else if (i %in% c(140:239,780,781,784,790:799,240:249,251:279,680:709,782)) return('Neoplasms')
  else return('other')
} # the fac function is not a vector, need to Vectorize it.

Diadata2$diag_1 = as.numeric(Diadata2$diag_1) #convert diag_1 to numberic data.
Diadata2 = Diadata2[complete.cases(Diadata2),]


test=mutate(Diadata2,prim_diag=sapply(Diadata2$diag_1, fac)) 

test$prim_diag <- factor(test$prim_diag, 
                         levels = c("Circulatory","Respiratory","Neoplasms",
                                    "Digestive","other","Diabetes","Injury",
                                    "Musculoskeletal", "Genitourinary"))

count <- test %>%
  group_by(prim_diag) %>%
  summarise(.,count = n())
A=inner_join(test,count,by = 'prim_diag')
############################################
rename_A1Cresult = function(i,j) {
  if (i == ">8" & j == "No")  return('High_medical not change')
  else if (i == ">8" & j == "Ch") return('High_medical change')
  else if(i==">7" & j == "No") return('Low_medical not change')
  else if (i == ">7" & j == "Ch") return('Low_medical change')
  else if(i == "Norm") return("Norm")
  else return('None')
}
exam <- filter(test, A1Cresult !=">7") %>%
  mutate(.,new_A1Cresult=mapply(rename_A1Cresult, .$A1Cresult,.$change)) 

#arrange the new_A1Cresult by ("None", "Norm", "High_medical change", "High_medical not change"):

exam$new_A1Cresult = factor(exam$new_A1Cresult, levels = c("None", "Norm", "High_medical change", "High_medical not change"))

test$max_glu_serum = factor(test$max_glu_serum, levels = c("None", "Norm",">200",">300"))
test$A1Cresult = factor(test$A1Cresult, levels = c("None", "Norm",">7",">8"))

# With all the case, including A1Crest !>"7". We find the result for A1Cresult is quite small and not a obvious difference.
#exam <- mutate(test,new_A1Cresult=mapply(rename_A1Cresult, test$A1Cresult,test$change)) 

###################################################
#calcute the ratio for readmittted for each case.

# count the number of each primary diagnosis:
count_A1C_primdiag <-  exam  %>%
  group_by(., new_A1Cresult, prim_diag)  %>%
  summarise(., total_A1C_primdiag = n())
# count the number of readmitted < 30days for each primary diagnosis:
count_A1C_primdiag_readmitted <-  exam  %>%
  group_by(., new_A1Cresult,prim_diag, readmitted)  %>%
  summarise(., total_A1C_primdiag_admitted30 = n()) %>%
  filter(., readmitted == "<30") %>%
  select(.,total_A1C_primdiag_admitted30)
# Generate a new data frame for ratio of readmitted people number vs each primary diagnosis people number:
ratio_readmitted <- inner_join(count_A1C_primdiag,count_A1C_primdiag_readmitted, by=c("new_A1Cresult","prim_diag")) %>%
  mutate(., ratio = total_A1C_primdiag_admitted30/total_A1C_primdiag)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Outlines

- The primarily diagnosed diseases and populations of diabetic patients
- The diabetic population with HbA1c test
- The impact of HbA1c test result on medication number for each primary diagnosis
- The impact of HbA1c test result on hospital readmission ratio for each primary diagnosis
- Summary





## Slide with R Output

```{r diabetic_data, echo = FALSE}
dim(Diadata)
dim(test)
names(test)[1:15]
names(test)[19]
names(test)[48:51]
```

## The diabetic population for each primary diagnosis 

```{r }


#####################################

ggplot(A, aes(x=reorder(prim_diag,count)))+
  geom_bar(fill="red3",size=1, width = 0.7)+
  coord_flip() +
  ggtitle("Primary Diagnosis in Final Dataset")+
  xlab("Primary Diagnosis") +
  theme(plot.title = element_text(lineheight=.8, face="bold"),
        axis.text=element_text(size=12),
        axis.title=element_text(size=14,face="bold"),
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "grey0",size = 0.1, linetype = 3),
        panel.grid.minor = element_blank())


```

##  The diabetic population with HbA1c testing

```{r pressure}

ggplot(exam, aes(x=new_A1Cresult))+
  geom_bar(fill="green4",size=1, width = 0.7)+
  scale_x_discrete(labels = c("None" = "None","Norm"="Normal",
                              "High_medical change" ="High\n Medical\nChange",
                              "High_medical not change" = "High\n Medical\n No Change")) +
  xlab("HbA1c Test")+
   ylab("Population ") +
   theme(
        axis.text = element_text(size = 7),
        plot.title = element_text(size=11, vjust = 2, face="bold"),
        axis.title = element_text(size=9,face="bold"))

#ggsave("age_with_race.pdf")
```

## HbA1c affects medication number

```{r }

################################################
ggplot(exam,aes(x=new_A1Cresult,y = num_medications))+
  geom_boxplot()+
  ylim(c(1,40))+
  facet_wrap(~prim_diag)+
  ggtitle("Patients' Medication Numbers VS HbA1C Test Results") +
  xlab("HbA1c Test")+
  ylab("Medication Number ")+
  scale_x_discrete(labels = c("None" = "None","Norm"="Normal",
                              "High_medical change" ="High\n Medical\nChange",
                              "High_medical not change" = "High\n Medical\n No Change")) +
  theme(legend.position = "none",
        axis.text = element_text(size = 7),
        plot.title = element_text(size=11, vjust = 2, face="bold"),
        axis.title = element_text(size=9,face="bold"),
        panel.margin = unit(0.1,"in"),
        strip.text = element_text(face = c("italic","bold"), size = 9, colour= "blue3")
  ) 
#ggsave("HbA1c_vs_numbermedication.pdf")
```

## HbA1c affects readmittted ratio

```{r }


# Plot the readmitted ratio for each promary diagnosis based on A1Cresult.

pp  <-   ggplot(ratio_readmitted, aes(x= new_A1Cresult, y=ratio,fill=new_A1Cresult)) +
  geom_bar(stat = "identity", width = 0.5)+
  facet_wrap(~ prim_diag)+
  ggtitle("Patients' Hospital Readmission Ratios VS HbA1C Test Results") +
  xlab("HbA1c Test")+
  ylab("Readmitted Ratio")+
  scale_x_discrete(labels = c("None" = "None","Norm"="Normal",
                              "High_medical change" ="High\n       Medical\nChange",
                              "High_medical not change" = "High\n Medical\n No Change")) +
  theme(legend.position = "none",
        axis.text = element_text(size = 7),
        plot.title = element_text(size=11, vjust = 2, face="bold"),
        axis.title = element_text(size=9,face="bold"),
        panel.margin = unit(0.1,"in"),
        strip.text = element_text(face = c("italic","bold"), size = 9, colour= "blue3")
  ) 

pp
#The plot suggests that the relationship between
#the probability of readmission and the HbA1c measurement 
#signicantly depends on the primary diagnosis 
#note that diabetes is always one of the secondary diagnoses

#ggsave("readmission ratio for HbA1C.pdf")

```

## Summary

- Diabetic patients are primarily diagnosed as serval diseases. Most of the diagnoses are performed without HbA1c test. However, HbA1c test is a very improtant parameter for efficient treatment of diabetic patients.  

- Using the mdication numbers of normal HbA1c result as baseline, the high H1bAc test with medication change results in an increase medication number. However, that without medication change reduces medication number.

- The relationship between the readmission ratio and the HbA1c measurement signicantly depends on the primary diagnosis. Our analyses indicate whether medication change of high HbA1c result is effective for different primary diagnosis.  
